Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-13 talk/pepm2011/paper.tex
Try to solve the XXXs of Michael. Add a new figure with the operation numbers.

\newcommand\cfbolz[1]{\nb{CFB}{#1}}
\newcommand\arigo[1]{\nb{AR}{#1}}
\newcommand\fijal[1]{\nb{FIJAL}{#1}}
\newcommand\david[1]{\nb{DAVID}{#1}}
\newcommand\anto[1]{\nb{ANTO}{#1}}
\newcommand\reva[1]{\nb{Reviewer 1}{#1}}
\newcommand\revb[1]{\nb{Reviewer 2}{#1}}
\newcommand\revc[1]{\nb{Reviewer 3}{#1}}
\newcommand{\commentout}[1]{}

@@

Moreover, thanks to the internal design of the JIT generator, it is very easy
to add new \emph{backends} for producing the actual machine code, in addition
to the original backend for the Intel \emph{x86} architecture.  Examples of
additional JIT backends are the one for Intel \emph{x86-64} and an
experimental one for the CLI .NET Virtual Machine \cite{cuni_python_cli_2010}.
PyPy's JIT generator generates a \emph{tracing JIT
compiler}, a concept which we now explain in more details.

@@

This trace of operations is then the basis of the generated code. The trace is
first optimized, and then turned into machine code. Both optimizations
and machine code generation is simple, because the traces are linear. This
linearity makes many optimizations a lot more tractable, and the inlining that
happens gives the optimizations automatically more context to work with.

@@

\newcommand{\ignore}[1]{} % {{\tt \small ignore(#1)}}
\ignore{
class Base(object):
    def add(self, other):
        """ add self to other """
        raise NotImplementedError("abstract base")
    def add__int(self, intother):
        """ add intother to self,
            where intother is an integer """
        raise NotImplementedError("abstract base")
    def add__float(self, floatother):
        """ add floatother to self,
            where floatother is a float """
        raise NotImplementedError("abstract base")
    def is_positive(self):
        """ returns whether self is positive """
        raise NotImplementedError("abstract base")
        }
\begin{figure}
\begin{lstlisting}[mathescape]
class Base(object):
      ...

@@

def f(y):
   res = BoxedInteger(0)
   while y.is_positive():
      res = res.add(y).add(BoxedInteger(-100))
      y = y.add(BoxedInteger(-1))
   return res
\end{lstlisting}
\caption{An ``interpreter'' for a tiny Dynamic Language written in RPython}
%\caption{A Simple Object Model and an Example Function Using it}
\label{fig:objmodel}
\end{figure}

@@

XXX this is not an RPython interpreter; put a reference to the previous
paper to show how we deal with an interpreted piece of code and remove
the interpretation overhead, turning it into basically something
equivalent to the example here, which is the start of the present paper.

@@

XXX in which language is the trace written in ? still RPython ?

@@

XXX explain set and get + int\_add briefly

The trace shows the inefficiencies of \lstinline{f} clearly, if one
looks at the number of \lstinline{new} (corresponding to object creation),
\lstinline{set/get} (corresponding to attribute reads/writes) and
\lstinline{guard_class} operations (corresponding to method calls).
In the rest of the paper we will see how this trace can be optimized using
partial evaluation.

@@

@@

This is a process that is usually called \emph{escape analysis}. In this paper we will
perform escape analysis by using partial evaluation. The partial evalution is a
bit peculiar in that there are not actually any constant arguments to the trace,
but it is only used to optimized operations within a trace. XXX mention Prolog.

@@

\begin{figure*}
\begin{tabular}{lrrrrrrrrrrrrrrrrrrrrrrr}
	&\#loops	&new	&	&	&get	&	&	&set	&	&	&guard	&	&	&rest	&	&	&	&	&	&total	&	&\\
crypto\_pyaes	&75	&1663	&446	&27\%	&18842	&2127	&11\%	&14552	&2020	&14\%	&8952	&235	&3\%	&53339	&7794	&15\%	&44009	&4828	&11\%	&97348	&12622	&13\%\\
django	&53	&453	&155	&34\%	&11970	&1457	&12\%	&8738	&991	&11\%	&4093	&275	&7\%	&33988	&4076	&12\%	&25254	&2878	&11\%	&59242	&6954	&12\%\\
fannkuch	&40	&161	&76	&47\%	&574	&306	&53\%	&278	&139	&50\%	&1121	&202	&18\%	&2581	&1685	&65\%	&2134	&723	&34\%	&4715	&2408	&51\%\\
go	&510	&5604	&1651	&29\%	&112403	&12612	&11\%	&88439	&11915	&13\%	&51621	&3194	&6\%	&300831	&30834	&10\%	&258067	&29372	&11\%	&558898	&60206	&11\%\\
html5lib	&416	&5939	&1793	&30\%	&241402	&15863	&7\%	&178723	&13960	&8\%	&58486	&3037	&5\%	&682654	&40341	&6\%	&484550	&34653	&7\%	&1167204	&74994	&6\%\\
meteor-contest	&56	&221	&141	&64\%	&2326	&606	&26\%	&2047	&645	&32\%	&1040	&169	&16\%	&7018	&2250	&32\%	&5634	&1561	&28\%	&12652	&3811	&30\%\\
nbody	&10	&84	&48	&57\%	&287	&159	&55\%	&121	&74	&61\%	&411	&90	&22\%	&986	&737	&75\%	&903	&371	&41\%	&1889	&1108	&59\%\\
pyflate-fast	&161	&1878	&723	&38\%	&22815	&2592	&11\%	&18415	&3690	&20\%	&8291	&628	&8\%	&64649	&11327	&18\%	&51399	&7633	&15\%	&116048	&18960	&16\%\\
raytrace-simple	&117	&1906	&359	&19\%	&52995	&2202	&4\%	&39518	&1823	&5\%	&13451	&447	&3\%	&140723	&6533	&5\%	&107870	&4831	&4\%	&248593	&11364	&5\%\\
richards	&79	&375	&191	&51\%	&27507	&2095	&8\%	&21217	&2004	&9\%	&3972	&332	&8\%	&76751	&6171	&8\%	&53071	&4622	&9\%	&129822	&10793	&8\%\\
spambayes	&267	&2413	&505	&21\%	&53327	&3033	&6\%	&40938	&2428	&6\%	&19343	&718	&4\%	&143575	&10322	&7\%	&116021	&6684	&6\%	&259596	&17006	&7\%\\
spectral-norm	&35	&266	&88	&33\%	&3004	&398	&13\%	&2544	&414	&16\%	&1084	&78	&7\%	&4803	&1184	&25\%	&6898	&978	&14\%	&11701	&2162	&18\%\\
telco	&42	&660	&70	&11\%	&20716	&424	&2\%	&14444	&608	&4\%	&6574	&56	&1\%	&53674	&1224	&2\%	&42394	&1158	&3\%	&96068	&2382	&2\%\\
twisted-names	&201	&2706	&317	&12\%	&64438	&2795	&4\%	&47960	&1968	&4\%	&24091	&555	&2\%	&171582	&6828	&4\%	&139195	&5635	&4\%	&310777	&12463	&4\%\\
total	&2062	&24329	&6563	&27\%	&632606	&46669	&7\%	&477934	&42679	&9\%	&202530	&10016	&5\%	&1737154	&131306	&8\%	&1337399	&105927	&8\%	&3074553	&237233	&8\%\\
\end{tabular}
\caption{Number of Operations Before and After Optimization}
\label{fig:numops}
\end{figure*}

>>>>

\newcommand\cfbolz[1]{\nb{CFB}{#1}}
\newcommand\arigo[1]{\nb{AR}{#1}}
\newcommand\fijal[1]{\nb{FIJAL}{#1}}
\newcommand\david[1]{\nb{DAVID}{#1}}
\newcommand\anto[1]{\nb{ANTO}{#1}}
\newcommand\reva[1]{\nb{Reviewer 1}{#1}}
\newcommand\revb[1]{\nb{Reviewer 2}{#1}}
\newcommand\revc[1]{\nb{Reviewer 3}{#1}}
\newcommand{\commentout}[1]{}
\newcommand{\ignore}[1]{} % {{\tt \small ignore(#1)}}

@@

Moreover, thanks to the internal design of the JIT generator, it is very easy
to add new \emph{backends} for producing the actual machine code, in addition
to the original backend for the Intel \emph{x86} architecture.  Examples of
additional JIT backends are the one for Intel \emph{x86-64} and an
experimental one for the CLI .NET Virtual Machine \cite{cuni_high_2010}.
PyPy's JIT generator generates a \emph{tracing JIT
compiler}, a concept which we now explain in more details.

@@

This trace of operations is then the basis of the generated code. The trace is
first optimized, and then turned into machine code. Both optimization
and machine code generation is simple, because the traces are linear. This
linearity makes many optimizations a lot more tractable, and the inlining that
happens gives the optimizations automatically more context to work with.

@@

\begin{figure}
\begin{lstlisting}[mathescape]
class Base(object):
   pass

@@

def f(y):
   res = BoxedInteger(0)
   while y.is_positive():
      res = res.add(y).add(BoxedInteger(-100))
      y = y.add(BoxedInteger(-1))
   return res
\end{lstlisting}
\caption{An ``Interpreter'' for a Tiny Dynamic Language Written in RPython}
\label{fig:objmodel}
\end{figure}

@@

@@

@@

The operations in the trace correspond to the operations in the RPython program
in Figure~\ref{fig:objmodel}:

\begin{itemize}
    \item \lstinline{new} corresponds to object creation.
    \item \lstinline{get} correspond to attribute reads.
    \item \lstinline{set} correspond to attribute writes.
    \item \lstinline{guard\_class} correspond to method calls and are followed by
    the trace of the called method.
    \item \lstinline{int\_add} and \lstinline{int\_get} are integer addition and
    comparison (``greater than''), respectively.
\end{itemize}

@@

The trace shows the inefficiencies of \lstinline{f} clearly, if one looks at the
number of \lstinline{new}, \lstinline{set/get} and \lstinline{guard\_class} operations.
In the rest of the paper we will see how this trace can be optimized using
partial evaluation.

@@

This is a process that is usually called \emph{escape analysis}. In this paper we will
perform escape analysis by using partial evaluation. The partial evalution is a
bit peculiar in that there are not actually any constant arguments to the trace,
but it is only used to optimized operations within a trace.

@@

\begin{figure*}
\begin{tabular}{|l||r|rr|rr|rr|rr|}
\hline
	&num loops	&new	&removed	&get/set	&removed &guard &removed	&all ops	&removed\\
\hline
crypto\_pyaes	&78	&3088	&50\%	&57148	&25\%	&9055	&95\%	&137189	&80\%\\
django	&51	&673	&54\%	&19318	&18\%	&3876	&93\%	&55682	&85\%\\
fannkuch	&43	&171	&49\%	&886	&63\%	&1159	&81\%	&4935	&45\%\\
go	&517	&12234	&76\%	&200842	&21\%	&53138	&90\%	&568542	&84\%\\
html5lib	&498	&14432	&68\%	&503390	&11\%	&71592	&94\%	&1405780	&91\%\\
meteor-contest	&59	&277	&36\%	&4402	&31\%	&1078	&83\%	&12862	&68\%\\
nbody	&13	&96	&38\%	&443	&69\%	&449	&78\%	&2107	&38\%\\
pyflate-fast	&162	&2278	&55\%	&39126	&20\%	&8194	&92\%	&112857	&80\%\\
raytrace-simple	&120	&3118	&59\%	&91982	&15\%	&13572	&95\%	&247436	&89\%\\
richards	&87	&844	&4\%	&49875	&22\%	&4130	&91\%	&133898	&83\%\\
spambayes	&314	&5608	&79\%	&117002	&11\%	&25313	&94\%	&324125	&90\%\\
spectral-norm	&38	&360	&64\%	&5553	&20\%	&1122	&92\%	&11878	&77\%\\
telco	&46	&1257	&90\%	&37470	&3\%	&6644	&99\%	&98590	&97\%\\
twisted-names	&214	&5273	&84\%	&100010	&10\%	&23247	&96\%	&279667	&92\%\\
\hline
total	&2240	&49709	&70\%	&1227447	&14\%	&222569	&93\%	&3395548	&89\%\\
\hline
\end{tabular}
\caption{Number of Operations and Percentage Removed By Optimization}
\label{fig:numops}
\end{figure*}
